Normalizations are not what you think; Explicit Scale Simulation in ALDEx2

### 2023-10-21 go3.Rmd
submitted version

---
title: "Beyond compositionally in high throughput sequencing; estimating the importance of scale in data analysis with ALDEx2
"
shorttitle: "Scale ALDEx2"
author:
- name: Greg Gloor
  affiliation: Dep't of Biochemistry, University of Western Ontario
  email: ggloor@uwo.ca
- name: Michelle Pistner Nixon
  affiliation: College of Information Sciences and Technology, Pennsylvania  State University
- name: Justin Silverman
  affiliation: College of Information Sciences and Technology, Department of Statistics, Department of Medicine, Pennsylvania  State University
bibliography: /Users/ggloor/Library/texmf/bibtex/bib/bibdesk_refs.bib
output:
  BiocStyle::pdf_document: default
package: ALDEx2
abstract: |
    Introduction to scale simulation and FDR correction with ALDEx2.
---

@@

High throughput sequencing (HTS) is a universal tool to explore many biological phenomenon such as gene expression (single-cell sequencing, RNA-sequencing, meta-transcriptomics), microbial community composition (16S rRNA gene sequencing, shotgun metagenomics) and differential enzyme activity (selex, CRISPR killing). HTS proceeds by taking a fixed-number sample from the environment, making a library, multiplexing (merging) multiple libraries and applying a fixed number of molecules to the flow cell. In essence it is a poll of the environment that is mixed with other polls and then a poll of the mixture is taken. It should be clear that the total number of molecules sequenced is driven by the capacity of the instrument and not by the number of molecules in the sampled environment[@Lovell:2015;@Vandeputte:2017aa;@Props:2017aa].

Data that are generated by sequencing come from systems where scale is usually important and may be a confounding variable [@Lovell:2015]. For example, cells transformed by the cMyc oncogene have about 3 times the amount of mRNA and about twice the rRNA content than do non-transformed cells [@Nie:2012aa], and this dramatically skews transcriptome analysis [@Loven:2012aa]. In addition, wild-type and mutant strains of cell lines, yeast or bacteria often have different growth rates, which would affect our ability to identify truly differentially abundant genes [@Yoshikawa:2011aa]. As another example, the total bacterial load of the vaginal microbiome differs by 1-2 orders of magnitude in absolute abundance between the healthy and bacterial vaginosis states [@Zozaya:2010]; the composition is dramatically different as well [@Ravel:2010;@Hummelen:2010]. Thus, the full description of these systems includes both relative change (composition) and absolute abundance (scale) but current methods access only the compositional information.

More formally, if the true information in the  environment being sampled can be summarized by counts of features (genes, taxa, etc) in a matrix \( \mathbf{W} \), then the elements of the matrix can be decomposed into its composition (relative information, \( \mathbf{W}^{\parallel} \) ) and its scale (total sum, \( \mathbf{W}^{\perp} \) ); formally, \(\mathbf{W} = \mathbf{W}^{\parallel} \times \mathbf{W}^{\perp}\). If the matrix has \(N\) samples and \(D\) parts, then we can uniquely identify \(\mathbf{W}_{dn}\) as the \(d^{th}\) feature in sample \(n\).  Any data that we collect by sequencing is thought to contain only proportional [@lovell:2011] information and we can represent the corresponding sequenced matrix as \(\mathbf{Y}\). Compositional data analysis makes the strong assumption  that \(\mathbf{Y}_{n} \equiv \mathbf{Y}^{\parallel}_{n} = f(\mathbf{W}^{\parallel}_{n})\); in other words that what is measured by sequencing is relative information only, and that this measurement is a function of the relative information in the underlying environment. Note that under this  assumption no corresponding estimate or information is available for the scale of the sample \(\mathbf{Y}^{\perp}_{n}\).

One issue that was realized very early was that if the goal was to compare one sample to another then the output from HTS needed to be normalized in order to make the samples commensurate and to correct any minor asymmetries in the data. A large number of normalizations were developed. Rarefaction[@Hughes:2005tu] was used in the microbiome field, while transcriptome investigators developed and used proportions and derivatives (reads per kilobase per million, and transcripts per kilobase per million[@Mortazavi:2008; @wagner:tpm]), the relative log expression (RLE)[@Anders:2010] and trimmed mean of M values (TMM)[@Robinson:2010a]. The centered log ratio (CLR)[@aitchison1982]  approach was proposed as a unifying normalization[@fernandes:2014]. With the exception of rarefaction, all of these normalizations are ratios with the major differences between approaches being how the denominator is chosen and whether the ratio is always log transformed or not.

Recently, Nixon et al. [-@nixon2023scale] showed that that *all* normalizations make some  identifying assumption about the scale of \(\mathbf{Y}\) but that the actual assumption being made was not always easy to interpret. The CLR which can be written as \( \mathbf{CLR}_{dn} = \mathbf{Y}_{dn} / G_n \) uses the geometric mean \(G_n\) of \(\mathbf{Y}^{\parallel}_{n}\) to scale the data. They discovered that \(G_n\) was related to the scale of the underlying environment \( \mathbf{W\perp} \). Interestingly, the information in a system is a fundamental property and \(G_n\) is strongly correlated with Shannon's entropy, indicating that information about the complexity of \( \mathbf{W} \) is available post-sequencing (Supplemental). The realization that all normalizations made assumptions of scale led to the understanding that all normalizations  introduced latent bias into analyses because the scaling factor has been assumed up to now to be a perfect point estimate of the scale of the underlying system [@nixon2023scale].

With this new insight Nixon et al. [-@nixon2023scale] devised methods to set up a scale model that explicitly acknowledges and accounts for the uncertainty of the scale. In the case of the CLR which uses the  \(G_n\) as the denominator, this entails adding uncertainty into  \(G_n\) such that the latent bias can be exposed, acknowledged and accounted for.

The ALDEx2 R package [@fernandes:2013] represents a general purpose toolbox for Bayesian estimation of HTS datasets. ALDEx2 was designed originally to convert the discrete point estimates in the count matrix obtained into continuous posterior distributions through Monte-Carlo sampling from the Dirichlet distribution. The posterior more accurately represents variation on a per-sample basis[@gloorAJS:2016] and this \(\mathbf{Y}^{\parallel}_{dn}\) can more closely approximate \(\mathbf{W}^{\parallel}_{dn}\).  Each Monte-Carlo replicate is normalized using the CLR and used to calculate summary and test statistics, and finally  expected values and confidence intervals are reported across the Monte-Carlo replicates. The CLR calculation can be modified trivially to include a full scale model of the data by incorporating uncertainty into \(G\) and so produce a posterior estimate of the scale. The scale uncertainty  can be included on a per-experiment, per-group or even a per-sample basis. Thus, ALDEx2 can now be used to generate a full posterior model of both \(\mathbf{Y}^{\parallel}_{dn}\) and \(\mathbf{Y}^{\perp}_{dn}\).

@@

The first dataset is a highly replicated yeast transcriptome where one condition is wild-type and the other has a snf-1 gene knockout[@Gierlinski:2015aa]. Yeast deficient for snf-1 grow more slowly and are sensitive to a variety of common agents that cause cell stress [@Yoshikawa:2011aa]. This dataset has been used to argue that only a small number of replicates need to be used to identity differentially abundant genes and that different tools should be used for datasets with different sample sizes because the tools have different intrinsic statistical power and Type 1 and Type 2 errors [@Schurch:2016aa]. This guidance runs counter to standard statistical practice where power is intrinsically linked to sample size [@Halsey:2015aa], yet the concept of sample-size independent power is entrenched in all fields that use HTS as an experimental readout. Increasing false precision with increasing sample size can be understood as the result of unacknowledged bias brought about because of false certainty as to the scale of the data. Thus, by adding scale uncertainty we can be more confident that the analysis is not converging on a precise but biased estimate of scale [@gustafson2015bayesian].

We start by examining the dispersion or variance of the data as counts and as the logarithm of the normalized counts as calculated by DESeq2 (log(RLE)) and by ALDEx2 (CLR). The actual counts of data derived from sequencing are overdispersed with the mean value being less than the variance [@Robinson:2010] as seen in Panel A of Figure 1. This relationship is why most tools model the counts with the Negative Binomial distribution[@polyester:2016], and use this distribution as the basis for batch correction[@Zhang:2020ab]. However, the actual analysis of differential abundance is performed on the logarithm of the normalized counts [@Robinson:2010; @Love:2014aa; @Andrews:2019aa], or on  the clr values [@fernandes:2013] both of which are log-ratios. The mean-dispersion distribution of these log-ratio transformed data is quite different as shown in panels B and C of Figure 1 and as noted elsewhere [@fernandes:2013; @Love:2014aa].

```{r dispersion, echo=F, warning=F, message=F,comment=F, result=F, fig.cap="Plot of abundance v dispersion for a typical transcriptome dataset as counts,  as logarithms of counts, and as clr values. Panel A shows that the data are over-dispersed relative to a Poisson distribution which is represented by the dashed line when plotted on a log-log scale. Panel B shows that the relationship between the mean and the dispersion calculated in DESeq2, here the standard error (SE) of the mean, is very different when the data are log-transformed first. Panel C shows the equivalent values calculated by ALDEx2  in which the expected clr value for each transcipt are plotted vs. the expected dispersion. The red line in each panel shows the loess line of fit to the mid-point of the distributions. In panels B and C the amount of dispersion reaches a minimum at moderate  values.  The dashed orange line in panel A is the line of equivalence, and in panel B and C is the minimum y value. The values below the dashed grey line in panels B and C represent those below the first decile of dispersion." }

library(DESeq2)
devtools::load_all('~/Documents/0_git/ALDEx_bioc')
#library(ALDEx2)

@@

par(mfrow=c(1,3))
plot(log2(yst.mn), log2(yst.v), pch=19, cex=0.5, col=rgb(0,0,0,0.2),
  xlab='Mean count', ylab='Variance')
#points(log2(yst.mn[sig.des]), log2(yst.v[sig.des]), pch=19, cex=0.5, col='orange')
def <- data.frame(log2(yst.mn),log2(yst.v))
lines(lowess(def, f=0.1), col=2, lwd=2)
title('A: Mean vs. Variance', adj=0, line= 0.8)
abline(0,1, col=rgb(1,0.6,0,0.5), lwd=2, lty=2)

l.df <- data.frame(yst.mn.log, res@listData$lfcSE)
plot(l.df, pch=19, cex=0.5, col=rgb(0,0,0,0.2),
  xlab='Mean log2(count)', ylab='SE log2(count)',ylim=c(0,0.4))
lines(lowess(l.df, f=0.25), col=2, lwd=2)
abline(h=min(lowess(l.df)$y), col=rgb(1,0.6,0,0.5), lwd=2, lty=2)
 abline(h=0.0275, col='grey', lty=2)
title('B: log2 Mean vs. log2 Variance', adj=0, line= 0.8)

l.df <- data.frame(x.all$rab.all, x.all$diff.win)
plot(x.all$rab.all, x.all$diff.win, cex=0.5, col=rgb(0,0,0,0.2),
  xlab='E(clr)', ylab='E(dispersion)')
#points(x.all$rab.all[sig.ald], x.all$diff.win[sig.ald], col='orange', cex=0.4)
lines(lowess(l.df, f=.1), col=2, lwd=2)
abline(h=min(lowess(l.df)$y), col=rgb(1,0.6,0,0.5), lwd=2, lty=2)
 abline(h=0.208, col='grey', lty=2)
title('C: E(clr) vs. E(dispersion)', adj=0, line= 0.8)
```

Here we can see that while the variance or dispersion always increases with increasing raw read count, it actually decreases when measured on the log-ratio transformed data [@fernandes:2013; @Love:2014aa] and reaches a minimum at some mid-point of the distribution. This makes the counter-intuitive suggestion that genes with moderate expression have more predictable expression than genes with very high expression such as housekeeping genes. This is at odds with the known biology of cells where single cell counting of housekeeping transcripts shows that they are both highly expressed and have little intrinsic variation [@Taniguchi:2010aa]. Furthermore, the actual amount of dispersion is very small and for many transcripts is almost negligible. To show this point more clearly, the majority of the transcripts in the lowest decile of dispersion indicated below the dashed grey line are statistically significantly different (75% with DESeq2, 69% with ALDEx2), suggesting that low dispersion estimates lead to many false positives because of the underestimation of dispersion. The work of Nixon et al. [-@nixon2023scale] suggests that this is because the biased scale estimates that are  calculated. Thus, the actual variation of highly expressed genes is not captured accurately by current approaches to analysis and supports the idea that HTS have unacknowledged bias.

Using either DESeq2 or ALDEx2, we observe that a majority of transcripts are statistically significantly different between groups even with a Benjamini-Hochberg [@benjamini:1995] false discovery rate (FDR) of 0.01; i.e. `r length(sig.des)` or `r length(sig.ald)` of the `r nrow(yst)` transcripts are statistically significant. Clearly, the necessary assumption of most features being invariant is not justified. Applying the rule of thumb of at least a $2^{1.4}$ fold change [@Schurch:2016aa] reduces these outputs to `r length(which(res@listData$padj < 0.01 & abs(res@listData$log2FoldChange) > 1.4))` for DESeq2 and to `r length(which(x.all$we.eBH < 0.01 & abs(x.all$diff.btw) > 1.4))` for ALDEx2.

@@

As shown using the effect  [@gloor:effect] and volcano plots [@Cui:2003aa] in Figure 2 A,B,D,E the root cause of the many statistically significant positive transcripts is the very large number of transcripts with negligible variance with both DESeq2 and ALDEx2. We can see that almost all the transcripts that are differentially abundant with an FDR < 0.01 (orange and red points) have extremely low dispersion and a very low difference between groups. In the most extreme cases transcripts with near 0 difference have a low FDR. This issue lead to the common practice of choosing transcripts with a low FDR and a fold-change threshold [@Schurch:2016aa] (commonly set at $\pm 2^{1.4}$), and these limits are shown by the dashed grey lines. A similar sitation arises when using the ALDEx2 package, and indeed the two methods identify substantially similar transcripts. This results begs the question, "why bother with the signficance test?".

The very low dispersion estimate for most of the features  arises because  scale variation in the underlying data has been removed through sequencing and normalization. While the actual scale of the data is inaccessible post-sequencing  we can estimate the *effect* of scale on the output. To do this, we add uncertainty to the denominator that is used to calculate the log-ratio of the samples [@nixon2023scale], and combine this with the posterior probabilities that ALDEx2 calculates on a per-feature basis [@fernandes:2013]. Scale uncertainty is incorporated using the `gamma` parameter that controls the amount uncertainty being included when we call either `aldex()`, or `aldex.clr()`. The `ALDEx2` package further contains a sensitivity analysis function, `aldex.senAnalysis()`, that can be used to explore the effect of different amounts of scale uncertainty. In practice we suggest that a `gamma` parameter between 0.5 and 1 is realistic for most experimental designs.

Applying `gamma=1` as a parameter we can see that the large number of transcripts with near 0 dispersion have had their dispersion increased (Figure 2C), and this results in many fewer transcripts (`r length(sig.s.ald)`) being called significantly different  as shown in the volcano plot in Figure 1F (red points). Furthermore, overplotting the significant transcripts identified after adding scale uncertainty on the un-scaled analysis shows that adding scale uncertainty removes the need for the dual cutoff.  Indeed, adding scale uncertainty reduces the significant transcripts to approximately the number observed with the somewhat arbitrary difference cutoff. Thus, incorporating scale uncertainty through the default scale model allows us to determine which variables are likely to be significant due to sequencing and normalization, and which are significantly different even with scale uncertainty included.

>>>>

---
title: "Normalizations are not what you think; Explicit Scale Simulation in ALDEx2
"
shorttitle: "Scale ALDEx2"
author:
  - name: Greg Gloor
    affiliation: Department of Biochemistry, University of Western Ontario
    email: ggloor@uwo.ca
  - name: Michelle Pistner Nixon
    affiliation: College of Information Sciences and Technology, Pennsylvania	 State University
  - name: Justin Silverman
    affiliation: College of Information Sciences and Technology and Department of Medicine, Pennsylvania	 State University
bibliography: /Users/ggloor/Library/texmf/bibtex/bib/bibdesk_refs.bib
output:
  nucleic-acids-research.csl: pdf_document
package: ALDEx2
abstract: In many high-throughput sequencing (HTS) studies, sample-to-sample variation  in sequencing depth is  not driven by  variation in the scale (e.g., total size, microbial load, or total gene  expression) of the underlying biological systems being measured but rather is driven by technical factors. Typically the technical variation  is addressed using some form of statistical normalization.  Normalizations are data or parameter transformations that remove unwanted  technical variation in the hopes of facilitating analyses sensitive to scale;  e.g., differential abundance and differential expression analyses. Recently we  showed that any normalization makes implicit assumptions about the unmeasured  system scale and that errors in these assumptions can lead to dramatic increases in  false positive and false negative rates. Here we describe updates to the ALDEx2 R  package that mitigate these problems by directly modeling uncertainty in the  unmeasured system scale through the use of a \textit{scale model}. Scale  models generalize the idea of normalizations and can be thought of as  explicitly modeling potential error in the chosen normalization. Beyond  enhancing the robustness of HTS analyses, the use of scale models within ALDEx2   enhances the transparency and reproducibility of analyses by making implicit  normalizing assumptions an explicit part of the model building process.
---

@@

High-throughput sequencing (HTS) is a ubiquitous tool used to explore many biological phenomenon such as gene expression (single-cell sequencing, RNA-sequencing, meta-transcriptomics), microbial community composition (16S rRNA gene sequencing, shotgun metagenomics) and differential enzyme activity (selex, CRISPR killing). HTS proceeds by taking a sample from the environment, making a library, multiplexing (merging) multiple libraries together, and then applying a sample of the multiplexed library to the flow cell. Each of these steps can be though of as a sampling step: a pool of DNA that could be measured is sampled and only that subsampled DNA is carried over to subsequent steps. Within each sampling step, the connection between the size of the sampled DNA pool and the scale (e.g., size, microbial load, or total gene expression) of the measured biological system is degraded or lost. Increasingly, researchers are turning to modified experimental protocols (e.g., DNA spike-ins or cell counting) in an attempt to recover the lost biological variation in scale; see for example [@Vandeputte:2017aa;@Props:2017aa]. However, these protocols often do not recover meaningful biological variation but instead recover scale variation at some intermediate step in the sample preparation protocol (e.g., at the stage where the DNA spike-in was added). In short, the disconnect between sample-to-sample variation in sequencing depth and biological variation in scale remains an outstanding challenge.

Biological variation in scale can represent an important unmeasured confounder in HTS analyses [@Lovell:2015]. For example, cells transformed by the cMyc oncogene have about 3 times the amount of mRNA and about twice the rRNA content than  non-transformed cells [@Nie:2012aa], and this dramatically skews transcriptome analysis [@Loven:2012aa]. In addition, wild-type and mutant strains of cell lines, yeast or bacteria often have different growth rates, which would affect our ability to identify truly differentially abundant genes [@Yoshikawa:2011aa]. As another example, the total bacterial load of the vaginal microbiome differs by 1-2 orders of magnitude in absolute abundance between the healthy and bacterial vaginosis states [@Zozaya:2010], and the composition between these states is dramatically different [@Ravel:2010;@Hummelen:2010]. Thus, a full description of these systems includes both relative change (composition) and absolute abundance (scale), yet current methods access only the compositional information.

To be more concrete, let \(\mathbf{Y}\) denote an \(D \times N\) matrix of sequence counts with elements \(\mathbf{Y}_{dn}\) denoting the number of measured DNA molecules mapping to feature \(d\) (e.g., a taxon or gene) in sample \(n\). Let \(\mathbf{W}\) also represent a \(D \times N\) matrix but with elements \(\mathbf{W}_{dn}\) representing the true (as opposed to measured) amount of class \(d\) in the biological system from which sample \(n\) was obtained. Succinctly, \(\mathbf{Y}\) denotes the observed HTS data whereas \(\mathbf{W}\) denotes the true counts in the system under study. We can think of \(\mathbf{W}\) as consisting of two parts, the scale \(\mathbf{W}^{\perp}\) (e.g., totals) and the composition \(\mathbf{W}^{\parallel}\) (i.e., proportions). That is, \(\mathbf{W}^{\perp}\) is a \(N\)-vector with elements \(\mathbf{W}^{\perp}_{n}=\sum_{d}\mathbf{W}_{dn}\) while \(\mathbf{W}^{\parallel}\) is a \(D \times N\) matrix with elements \(\mathbf{W}^{\parallel}_{dn}=\mathbf{W}_{dn}/\mathbf{W}^{\perp}_{n}\). Note that with these definitions \(\mathbf{W}\) can be written as the element-wise combination of scale and composition: \(\mathbf{W}_{dn}=\mathbf{W}^{\parallel}_{dn}\mathbf{W}^{\perp}_{n}\). The technical variation in sequencing depth (\(\mathbf{Y}^{\perp}_{n}=\sum_{d}\mathbf{Y}_{dn}\)) implies that observed data \(\mathbf{Y}\) provides us with information about the system composition \(\mathbf{W}^{\parallel}\) but little to no information in the system scale \(\mathbf{W}^{\perp}\) (Lovell et al. 2011).

Early on, researchers recognized the potential confounding effects of the non-biological variation in sequencing depth and approached the problem using various forms of \textit{normalization}. The idea of normalization is to remove this variation thus making the samples commensurate and to correct any minor asymmetries in the data. A large number of normalizations were developed. Rarefaction [@Hughes:2005tu] was used in the microbiome field, while transcriptome investigators developed and used proportions and derivatives (reads per kilobase per million, and transcripts per kilobase per million) [@Mortazavi:2008; @wagner:tpm], the relative log expression (RLE) [@Anders:2010] and trimmed mean of M values (TMM) [@Robinson:2010a]. The Centered Log Ratio (CLR) [@aitchison1982]  was proposed as a unifying normalization [@fernandes:2014]. With the exception of rarefaction, all of these normalizations are data transformations that can be stated as ratios of the form \(\hat{\mathbf{Y}}_{dn}=\mathbf{Y}_{dn}/f(\mathbf{Y})\); the major differences being how the denominator is chosen and whether the ratio is subsequently always log-transformed or not.

Recently, Nixon et al. [-@nixon2023scale] suggested that the challenge of non-biological variation in sequencing depth be viewed as a problem of partially-identified models. Through this lens, they showed that \emph{all} normalizations make some assumption about the system scale \(\mathbf{W}^{\perp}\) in order to identify the chosen model. Moreover, these assumptions are often implicit, difficult to interpret, and provided different outputs when applied to the same dataset [@Dillies:2013;@Weiss:2017aa]. Intuitively, normalizations in widespread use assume that either one sample can be chosen as a reference to which the others are scaled (TMM), or that different sub-parts of each sample maintain a constant scale across samples (RLE, CLR and derivatives). Further, Nixon et al. [-@nixon2023scale] showed  that the original ALDEx2 model made a strict assumption about scale through the CLR normalization. They showed that this assumption could be remedied since ALDEx2 permits scale to be easily incorporated into its framework. In this paper, we discuss these modifications and show how including scale uncertainty can greatly improve modeling and lead to more robust and reproducible results.

## Adding Scale Uncertainty in ALDEx2

The ALDEx2 R package [@fernandes:2013] is a general purpose toolbox for Bayesian modeling of HTS data. For brevity and concreteness, we discuss ALDEx2 in its simplest form as a tool for estimating Log-Fold-Changes (LFC) (e.g., differential abundance or differential expression analysis), but note that it can be used to fit more complex linear models. At a high-level, ALDEx2 involves three steps: 1) a Bayesian model is used to estimate \(\mathbf{W}^{\parallel}\) given observations \(\mathbf{Y}\); 2) a normalization is used to estimate \(\mathbf{W}\) given the estimate \(\mathbf{W}^{\parallel}\); 3) the \(\mathbf{W}\) estimates are used to estimate LFC as part of differential abundance (or expression) analyses. For more details on ALDEx2 see Fernandes et al. [-@fernandes:2013].

By default, ALDEx2 uses the CLR normalization which can be written as \(\log \mathbf{W}_{dn}=\log [\mathbf{W}^{\parallel}_{dn}/G_{n}]\) where \(G_{n}\) denotes the geometric mean of the vector \((W^{\parallel}_{1n}, \dots, W^{\parallel}_{Dn})\) (see Supplemental Materials for how the CLR connects to Shannon entropy). Nixon et al. (2023) showed that in this context the CLR normalization equates to an implicit assumption that \(\mathbf{W}^{\perp}_{n}=1/G_{n}\). They showed that even slight errors in this assumption introduced bias into LFC estimate that was not accounted for in estimates of uncertainty (e.g., confidence intervals or credible sets). In fact, they showed that the only way in which the ALDEx2 model, or any model that uses  normalizations, could ever be calibrated (e.g., control Type-I Error rates) was if this assumption was exactly true. While this assumption may be approximately correct in some instances it is surely never exactly true. In support of this assertion it is known that false discovery rates vary widely by analysis and normalization method [@hawinkel2017;@Li:2022aa]. As a result, Nixon et al. (2023) suggested that models should consider potential error in the assumptions implied by normalizations.

Nixon et al. (2023) proposed the concept of a \textit{scale model} as a generalization of normalizations that account for potential error. They showed how scale models can be incorporated into ALDEx2, turning the ALDEx2 model into a specialized type of statistical model which they called a \textit{Scale Simulation Random   Variable} (SSRV). Recall that \(\mathbf{W}_{dn}=\mathbf{W}^{\parallel}_{dn}\mathbf{W}^{\perp}_{n}\).  As the HTS data only informs directly on $\mathbf{W}^{\parallel}_{dn}$, they proposed to augment analyses with a model for $\mathbf{W}^{\perp}_{n}$ (the scale model). Since the CLR normalization makes the assumption \(\mathbf{W}^{\perp}_{n}=1/G_{n}\), the CLR normalization can be generalized by considering probability models for the scale \(\mathbf{W}^{\perp}_{n}\) that have mean \(1/G_{n}\). For example, the following scale model generalizes the CLR:

\[\log \mathbf{W}^{\perp}_{n} = -\log G_{n} + \Lambda x_{n} \qquad \Lambda \sim N(0, \gamma^{2})\]

where \(\gamma\) is a tunable parameter that controls the degree of uncertainty in the CLR assumption and \(x_{n}\) denotes a binary condition indicator (e.g., \(x_{n}=1\) denotes case and \(x_{n}=0\) denotes control). Beyond theoretical studies, they demonstrated through the analysis of simulated and real data the remarkable performance improvements that could result from incorporating this type of scale uncertainty into analyses. Subsequently, we have made those modifications a permanent fixture of ALDEx2 which now represents the first software package designed for SSRV-based inference.

Beyond concerns of fidelity and rigor, scale models also enhance the reproducibility and transparency of HTS analyses. This occurs because rather than using an implicit assumption (e.g., \(\log \mathbf{W}^{\perp}=1/G_{n}\) for the CLR normalization), scale models make this an explicit part of the model building process. While it is beyond the scope of the present article, we note that there are many ways of building scale models that enhance the interpretability of the parameters and assumptions. Moreover, the scale model listed above, which is now the default in ALDEx2, also has an appealing interpretation. A detailed description of these points as well as the mathematical underpinnings of the updated ALDEx2 R package is forthcoming and will be made available as a companion publication. Here, we simply treat the parameter \(\gamma\) in the default scale model as an abstract term that represents uncertainty in the CLR assumption and note that larger values of $\gamma$ correspond to more uncertainty in this assumption. We also show that the ratio of scales between conditions can be used to build a full scale model when the conditions have dramatically different underlying scales.

@@

## Adding scale uncertainty replaces the need for dual significance cutoffs.

Gierliński et al. [-@Gierlinski:2015aa] conducted a highly replicated yeast transcriptome experiment comparing a wild-type strain with a snf1 gene knockout. This dataset has been used to argue that only a small number of replicates are needed to identity differentially abundant genes and makes the recommendation that some tools should be used in some conditions because each tool has different intrinsic statistical power and Type 1 and Type 2 errors [@Schurch:2016aa]]. This guidance runs counter to standard statistical practice where power is intrinsically linked to sample size [@Halsey:2015aa], yet the concept of sample-size independent power is entrenched in all fields that use HTS as an experimental readout. Through the lens of this paper, this is a hallmark of unacknowledged bias: a false certainty in the scale of the data brings about an increasing false precision with increasing sample sizes. Furthermore, the yeast transcriptome experiment has an unacknowledged difference in scale between conditions: yeast deficient for snf1 are smaller, grow more slowly and are sensitive to a variety of common agents that cause cell stress [@Yoshikawa:2011aa]].  Thus, by adding scale uncertainty, we can be more confident that the results of the analysis reflect true biological differences and not differences due to the assumptions made by the normalization method.

We start by examining the dispersion (variance) of the data as counts and as the logarithm of the normalized counts when calculated by DESeq2 (log(RLE)) or by ALDEx2 (CLR). The raw data counts derived from sequencing are overdispersed with the mean value being less than the variance [@Robinson:2010a;@Robinson:2010] as seen in Panel A of Figure 1. However, the actual analysis of differential abundance is performed on the logarithm of the normalized counts [@Robinson:2010;@Love:2014aa;@Andrews:2019aa], which includes the CLR [@fernandes:2013] yet the mean-dispersion distribution of these log-ratio transformed data is quite different as shown in panels B and C of Figure 1 and as noted elsewhere [@fernandes:2013;@Love:2014aa].

```{r, echo=F, warning=F, message=F,comment=F, result=F} 
library(DESeq2)
devtools::load_all('~/Documents/0_git/ALDEx_bioc')
#library(ALDEx2)
```
```{r dispersion, echo=F, warning=F, message=F,comment=F, result=F, cache=T, fig.cap="Plot of abundance v dispersion for a typical transcriptome dataset as counts,	 as logarithms of counts, and as clr values. Panel A shows that the data are over-dispersed relative to a Poisson distribution which is represented by the dashed line when plotted on a log-log scale. Panel B shows that the relationship between the mean and the dispersion calculated in DESeq2, here the standard error (SE) of the mean, is very different when the data are log-transformed first. Panel C shows the equivalent values calculated by ALDEx2	 in which the expected clr value for each transcipt are plotted vs. the expected dispersion. The red line in each panel shows the loess line of fit to the mid-point of the distributions. In panels B and C the amount of dispersion reaches a minimum at moderate	 values.  The dashed orange line in panel A is the line of equivalence, and in panel B and C is the minimum y value. The values below the dashed grey line in panels B and C represent those below the first decile of dispersion." }

@@

par(mfrow=c(1,3))
plot(log2(yst.mn), log2(yst.v), pch=19, cex=0.5, col=rgb(0,0,0,0.2),
  xlab='Mean count', ylab='Variance')
#points(log2(yst.mn[sig.des]), log2(yst.v[sig.des]), pch=19, cex=0.5, col='orange')
def <- data.frame(log2(yst.mn),log2(yst.v))
lines(lowess(def, f=0.1), col=2, lwd=2)
title('A: M v. V', adj=0, line= 0.8)
abline(0,1, col=rgb(1,0.6,0,0.5), lwd=2, lty=2)

l.df <- data.frame(yst.mn.log, res@listData$lfcSE)
plot(l.df, pch=19, cex=0.5, col=rgb(0,0,0,0.2),
  xlab='Mean log2(count)', ylab='SE log2(count)',ylim=c(0,0.4))
lines(lowess(l.df, f=0.25), col=2, lwd=2)
abline(h=min(lowess(l.df)$y), col=rgb(1,0.6,0,0.5), lwd=2, lty=2)
 abline(h=0.0275, col='grey', lty=2)
title('B: log2(M) v. log2(V)', adj=0, line= 0.8)

l.df <- data.frame(x.all$rab.all, x.all$diff.win)
plot(x.all$rab.all, x.all$diff.win, cex=0.5, col=rgb(0,0,0,0.2),
  xlab='E(clr)', ylab='E(disp)')
#points(x.all$rab.all[sig.ald], x.all$diff.win[sig.ald], col='orange', cex=0.4)
lines(lowess(l.df, f=.1), col=2, lwd=2)
abline(h=min(lowess(l.df)$y), col=rgb(1,0.6,0,0.5), lwd=2, lty=2)
 abline(h=0.208, col='grey', lty=2)
title('C: E(clr) vs. E(dispersion)', adj=0, line= 0.8)
```

These relationships depend on what is being analyzed: the variance or dispersion always increases with increasing raw (or normalzied) read count but decreases when measured on the log-ratio transformed data [@fernandes:2013;@Love:2014aa], reaching a minimum at some mid-point of the distribution. This makes the counter-intuitive suggestion that genes with moderate expression have more predictable expression than genes with very high expression such as housekeeping genes. This is at odds with the known biology of cells where single cell counting of housekeeping transcripts shows that they are both highly expressed and have little intrinsic variation [@Taniguchi:2010aa]. Furthermore, the dispersion is exceeding small being, for many transcripts, almost negligible. To show this point more clearly, the majority of the transcripts in the lowest decile of dispersion indicated below the dashed grey line are statistically significantly different (75\% with DESeq2, 69\% with ALDEx2), suggesting that low dispersion estimates lead to many false positives. From the perspective of Nixon et al. [-@nixon2023scale], it is reasonable to conclude that some of these results may be due to the choice of normalization. Indeed, benchmarking and comparison studies have repeatedly shown that the choice of normalization plays a role in which results are returned as significant; see for example [@maza2013], [@Dillies:2013] and [@Weiss:2017aa].

Using either DESeq2 or ALDEx2,  a majority of transcripts are statistically significantly different between groups even with a Benjamini-Hochberg [@benjamini:1995] false discovery rate (FDR) of 0.01; i.e. `r length(sig.des)` or `r length(sig.ald)` of the `r nrow(yst)` transcripts are statistically significant. Clearly, a 70\% positive rate seems biologically unjustified, and furthermore this breaks the necessary assumption made by the normalization methods that most features must be invariant. Finally, that `r length(setdiff(sig.ald, sig.des))` transcripts are identified by ALDEx2 and not DESeq2, while DESeq2 identifies `r length(setdiff(sig.des, sig.ald))` transcripts that ALDEx2 does not suggests that the choice of normalization plays a role in which results are returned as significant.

@@

As shown using effect [@gloor:effect] and volcano plots [@Cui:2003aa] in Figure 2 A,B,D,E, the root cause of the many statistically significant positive transcripts is the very large number of transcripts with negligible variance. This phenomenon is shared between methods: almost all the transcripts that are differentially abundant with an FDR \textless 0.01 (orange and red points) have extremely low dispersion and a very low difference between groups. In the most extreme cases, some transcripts have a near zero difference between groups. This issue is not unique to this data set and it is common practice to use a dual-cutoff by choosing transcripts based on a thresholds for both corrected p-values and fold-changes  [@Schurch:2016aa] (commonly set at \(\pm 2^{1.4}\) for the latter). These limits are shown by the dashed grey lines. In this case, applying a dual-cutoff using a heuristic of at least a \(2^{1.4}\) fold change  reduces the number of significant outputs to 193 for DESeq2 and to 186 for ALDEx2. This result begs the question: why bother with significance tests at all?

The very low dispersion estimate for most features is a by-product of sequencing and normalization. While the actual scale of the data is inaccessible post-sequencing, we can study the sensitivity of the results to the choice of normalization and thus scale. To do this we incorporate scale models [@nixon2023scale] in lieu of normalizations through the default scale model that is now built into ALDEx2. Scale uncertainty is incorporated using the \texttt{gamma} parameter that controls the amount uncertainty added to the CLR mean assumption when we call either \texttt{aldex()}, or \texttt{aldex.clr()}. The \texttt{ALDEx2} package further contains a sensitivity analysis function, \texttt{aldex.senAnalysis()}, that can be used to explore the effect of different amounts of scale uncertainty. In practice, we suggest that a \texttt{gamma} parameter between 0.5 and 1 is realistic for most experimental designs.

Applying the default scale model with \texttt{gamma=1}, a large number of transcripts with near 0 dispersion have had their dispersion increased (Figure 2C), and this results in many fewer transcripts `r length(sig.s.ald)` being called significantly different as shown in the volcano plot in Figure 1F (red points). Furthermore, overplotting the significant transcripts identified after adding scale uncertainty on the un-scaled analysis shows that adding scale uncertainty removes the need for the dual cutoff. Indeed, adding scale uncertainty reduces the significant transcripts to a subset of those identified with the dual cutoff that have the largest effect size.  Thus, incorporating scale uncertainty through the default scale model allows us to determine which variables are likely to be significant due to sequencing and normalization, and which are significantly different even with scale uncertainty included.

## Housekeeping genes can be used to guide scale model choices.
